Alone Together: Compositional Reasoning and Inference for Weak Isolation

### 2017-10-22 inference.tex
Minor edits.

quantification that binds a set $x$ satisfying proposition $\phi$ in a

@@

The state transformer rules, like the earlier RG rules, closely follow
the corresponding reduction rules in Fig.~\ref{fig:txnimp}, except
that their language of expression is $\SL$. For instance, while the
reduction rule for \C{UPDATE} declaratively specifies the set of
updated records, the state transformer rule uses $\SL$'s bind
operation to \emph{compute} the set. Other SQL rules do likewise. The
rules for \C{LET} binders, conditionals, and sequences compose the
effects inferred for their subcommands. Thus, the effect of a sequence
of commands $c_1;c_2$ is the union of effects $F_1$ and $F_2$ of $c_1$
and $c_2$, respectively, except that $F_2$ is applied to a state
($\stl$) to which $F_1$ has already been applied, reflecting their
order of reduction. The inference rule for \C{FOREACH} takes advantage
of the $\SL$'s bind operator to lift the effect inferred for the loop
body to the level of the loop. Since records added to $\stl$ in each
iteration of \C{FOREACH} are independent of the previous iteration,
sequential composition of the effects of different iterations is same
as their parallel composition. Since the loop body is executed once
per each $z\in x$, the effect of the the loop is the union of effects
($\F$) for all $z\in x$, all applied to the same state ($\stl$ and
$\stg$). That is, $\F_{loop}(\stl,\stg) = \bigcup_{z\in
  x}\F_{body}(\stl,\stg)$. From the definition of the set monad's bind
operator, $\F_{loop}(\stl,\stg) = x \bind (\lambda
z.~F_{body}(\stl,\stg))$, which mirrors the definition of the rule.
This development mirrors ideas found in type checkers for certain
dependent type systems~\cite{KJ14}.

>>>>

quantification that binds a set $\stg$ satisfying proposition $\phi$ in a

@@

The state transformer rules, like the earlier RG rules, closely follow
the corresponding reduction rules in Fig.~\ref{fig:txnimp}, except
that their language of expression is $\SL$. For instance, while the
reduction rule for \C{UPDATE} declaratively specifies the set of
updated records, the state transformer rule uses $\SL$'s bind
operation to \emph{compute} the set. Other SQL rules do likewise. The
rules for \C{LET} binders, conditionals, and sequences compose the
effects inferred for their subcommands. Thus, the effect of a sequence
of commands $c_1;c_2$ is the union of effects $F_1$ and $F_2$ of $c_1$
and $c_2$, respectively, except that $F_2$ is applied to a state
($\stl$) to which $F_1$ has already been applied, reflecting their
order of reduction. The inference rule for \C{FOREACH} takes advantage
of the $\SL$'s bind operator to lift the effect inferred for the loop
body to the level of the loop. Since records added to $\stl$ in each
iteration of \C{FOREACH} are independent of the previous iteration
(recall local context independence; Sec. ~\ref{sec:opsem}), sequential
composition of the effects of different iterations is same as their
parallel composition. Since the loop body is executed once per each
$z\in x$, the effect of the the loop is the union of effects ($\F$)
for all $z\in x$, all applied to the same state ($\stl$ and $\stg$).
That is, $\F_{loop}(\stl,\stg) = \bigcup_{z\in
x}\F_{body}(\stl,\stg)$. From the definition of the set monad's bind
operator, $\F_{loop}(\stl,\stg) = x \bind (\lambda
z.~F_{body}(\stl,\stg))$, which mirrors the definition of the rule.
This development mirrors ideas found in type checkers for certain
dependent type systems~\cite{KJ14}.

### 2017-10-22 introduction.tex
Minor edits.

In this paper, we propose a program logic for weakly-isolated
transactions along with automated verification support to allow
developers to verify the soundness of their applications, without
having to resort to low-level operational reasoning as they are forced
to do currently.  We develop a set of syntax-directed compositional
proof rules that enable the construction of correctness proofs for
transactional programs in the presence of a weakly-isolated
concurrency control mechanism.  Realizing that the proof burden
imposed by these rules may discourage applications programmers from
using them, we also present an inference procedure based on these
rules that automatically verifies the weakest isolation level for a
transaction that still ensures its invariants are maintained.  The key
to inference is a novel formulation of database state (represented as
sets of tuples) as a monad, and in which database computations are
interpreted as state transformers over these sets.  This
interpretation leads to an encoding of database computations amenable
for verification by off-the-shelf SMT solvers.  The paper makes the
following contributions:
%% We have realized these ideas as an embedded DSL within OCaml, that
%% supports common relational SQL operations (updates, selects, inserts,
%% deletes, joins, etc.).  Experimental results on real-world benchmarks
%% demonstrate the feasibility and value of automated verification for
%% weakly isolated transactions, an important advance in improving the
%% safety of realistic database computations.
\begin{enumerate}
  \item We analyze properties of weak isolation in the context of a
    DSL embedded in OCaml that treats SQL-based relational database
    operations (e.g., inserts, selects, deletes, updates, etc.) as
    computations over an abstract database state.
  \item We develop an operational semantics and a compositional
    rely-guarantee style proof system for this language capable of
    relating high-level application invariants to database state,
    parameterized by a weak isolation semantics that selectively
    exposes the visibility of these operations to other transactions.
  \item We devise an inference algorithm capable of discovering the
    weakest isolation level that is sound with respect to a
    transaction's high-level consistency requirements. The algorithm
    interprets database operations as state transformers expressed in
    a language amenable for translation into a decidable fragment of
    first-order logic, and is thus suitable for automated verification
    using off-the-shelf SMT solvers.
  \item We present details of an implementation along with an
    evaluation study on real database benchmarks that justify our
    approach, and demonstrate the utility of our inference mechanism.
\end{enumerate}
\noindent Our results provide the first (to the best of our knowledge)
formalization of weakly-isolated transactions, along with an
expressive and compositional proof automation framework capable of
verifying the safety of high-level consistency conditions attached to
these transactions.  Collectively, these contributions allow
weakly-isolated transactions to enjoy the same rigorous reasoning
capabilities as their strongly-isolated (serializable) counterparts.

>>>>

In this paper, we propose a program logic for weakly-isolated
transactions along with automated verification support to allow
developers verify the soundness of their applications, without
having to resort to low-level operational reasoning as they are forced
to do currently.  We develop a set of syntax-directed compositional
proof rules that enable the construction of correctness proofs for
transactional programs in the presence of a weakly-isolated
concurrency control mechanism.  Realizing that the proof burden
imposed by these rules may discourage applications programmers from
using them, we also present an inference procedure based on these
rules that automatically verifies the weakest isolation level for a
transaction while ensuring its invariants are maintained.  The key to
inference is a novel formulation of database state (represented as
sets of tuples) as a monad, and in which database computations are
interpreted as state transformers over these sets.  This
interpretation leads to an encoding of database computations amenable
for verification by off-the-shelf SMT solvers.  The paper makes the
following contributions:
%% We have realized these ideas as an embedded DSL within OCaml, that
%% supports common relational SQL operations (updates, selects, inserts,
%% deletes, joins, etc.).  Experimental results on real-world benchmarks
%% demonstrate the feasibility and value of automated verification for
%% weakly isolated transactions, an important advance in improving the
%% safety of realistic database computations.
\begin{enumerate}
  \item We analyze properties of weak isolation in the context of a
    DSL embedded in OCaml that treats SQL-based relational database
    operations (e.g., inserts, selects, deletes, updates, etc.) as
    computations over an abstract database state.
  \item We develop an operational semantics and a compositional
    rely-guarantee style proof system for this language capable of
    relating high-level application invariants to database state,
    parameterized by a weak isolation semantics that selectively
    exposes the visibility of these operations to other transactions.
  \item We devise an inference algorithm capable of discovering the
    weakest isolation level that is sound with respect to a
    transaction's high-level consistency requirements. The algorithm
    interprets database operations as state transformers expressed in
    a language amenable for translation into a decidable fragment of
    first-order logic, and is thus suitable for automated verification
    using off-the-shelf SMT solvers.
  \item We present details of an implementation along with an
    evaluation study on real database benchmarks that justify our
    approach, and demonstrate the utility of our inference mechanism.
\end{enumerate}
\noindent Our results provide the first (to the best of our knowledge)
formalization of weakly-isolated transactions, along with an
expressive and compositional proof automation framework capable of
verifying the safety of high-level consistency conditions attached to
these transactions.  Collectively, these contributions allow
weakly-isolated transactions to enjoy the same rigorous reasoning
capabilities as their strongly-isolated (serializable) counterparts.

### 2017-10-22 motivation.tex
Minor edits.

\noindent This simple characterization of SI isolation allows us to
verify the consistency conditions associated with the \C{new\_order}
transaction.  First, since the database does not change ($\Delta' =
\Delta$) during execution of the transaction's body, we can reason
about \C{new\_order} as though it executed in complete isolation until
its commit point, leading to a verification process similar to what
would have been applied when reasoning sequentially.  When
\C{new\_order} finishes execution, however, but before it commits, the
SI axiomatization shown above requires us to consider global state
transitions $\stg \longrightarrow \stg'$ that do not include changes
to the records ($\stl$) written by \C{new\_order}, i.e.,
$\forall(x\in\mathit{dom}(\delta)).~\Delta'(x) = \Delta(x)$.  The
axiomatization precludes any execution in which there are concurrent
updates to shared table fields (e.g., \C{d\_next\_o\_id} on the same
\C{District} table), but does not prohibit interferences that write to
different tables, or write to different records in the same table.  We
need to reason about the safety of such interferences with respect to
\C{new\_order}'s consistency invariants.

@@

Fig.~\ref{fig:foreach_code} shows a (simplified) snippet of code taken
from Fig.~\ref{fig:new_order_code}. Some irrelevant expressions have
been replaced with constants (\C{k1},\C{k2}, and \C{k3}).  The body of
the loop executes a SQL update followed by an insert.  Recall that a
transaction reads from the global database ($\Delta$), and writes to a
transaction-local database ($\delta$) before committing these
updates. An update statement filters the records that match the search
criteria from $\Delta$ and computes the updated records that are to be
added to the local database. Thus, the state transformer for the
update statement (call it $\T_U$) is the following function on
sets\footnote{$\bind$ has higher precedence than $\cup$; angle braces
($\langle \ldots \rangle$) are used to denote records in math}:
\begin{smathpar}
\begin{array}{l}
  \lambda(\stl,\stg).~ \stl \cup \stg \bind(\lambda \C{s}. 
      \itel{{\sf table}(\C{s}) = \C{Stock} \conj 
        \C{s}.\C{s\_i\_id} = \C{item\_req.ol\_i\_id}\\\hspace*{1.15in}}
           {\{ \langle \C{s\_i\_id}=\C{s.s\_i\_id};\, 
                       \C{s\_d\_id}=\C{s.s\_d\_id};\,
                       \C{s\_qty} = \C{k1}\rangle \}\\\hspace*{1.15in}}
           {\emptyset})
\end{array}
\end{smathpar}
% \begin{smathpar}
% \begin{array}{c}
%   \lambda\Delta.\lambda\delta.\, \delta \;\cup\; 
%       \Delta \,\bind\, (\lambda s.\, 
%           \C{if}\;{\C{stock}(s) \conj 
%                    \C{s.s\_i\_id} = \C{item\_req.ol\_i\_id}}\\
%            \hspace*{0.9in}
%            \C{then}\;{\{\{s \;\C{with}\; \C{s\_qty} = \C{k1}\}\}}\;
%            \C{else}\;{\emptyset})
% \end{array}
% \end{smathpar}
% \begin{verbatim}
%   Rmem(σ'(s')) = Rmem(σ(s')) U 
%                  Rmem[\x. Rmem[\y. if y<x then {y+x} else {}](σ(s)) U {x}](xs)
% \end{verbatim}
% Observe that
% $\T_U(\Delta,\delta)$ is of the form $\delta \,\cup\,
% F_U(\Delta)$.
Here, the set bind operator extracts record elements ($\C{s}$) from
the database, checks the precondition of the update action, and if
satisfied, constructs a new set containing a single record that is
identical to $\C{s}$ except that it binds field $\C{s_{qty}}$ to value
$\C{k_1}$.  This new set is added (via set union) to the existing
local database state $\delta$.\footnote{For now, assume that the
  record being added to $\stl$ is not already present in $\stl$.}

>>>>

\noindent This simple characterization of SI isolation allows us to
verify the consistency conditions associated with the \C{new\_order}
transaction.  First, since the database does not change ($\Delta' =
\Delta$) during execution of the transaction's body, we can reason
about \C{new\_order} as though it executed in complete isolation until
its commit point, leading to a verification process similar to what
would have been applied when reasoning sequentially.  When
\C{new\_order} finishes execution, however, but before it commits, the
SI axiomatization shown above requires us to consider global state
transitions $\stg \longrightarrow \stg'$ that do not include changes
to the records ($\stl$) written by \C{new\_order}, i.e.,
$\forall(x\in\mathit{dom}(\delta)).~\Delta'(x) = \Delta(x)$.  The
axiomatization precludes any execution in which there are concurrent
updates to shared table fields (e.g., \C{d\_next\_o\_id} on the same
\C{District} table), but does not prohibit interferences that write to
different tables, or write to different records in the same table.  We
need to reason about the safety of such interferences with respect to
\C{new\_order}'s consistency invariants to verify \C{new\_order}.

@@

Fig.~\ref{fig:foreach_code} shows a (simplified) snippet of code taken
from Fig.~\ref{fig:new_order_code}. Some irrelevant expressions have
been replaced with constants (\C{k1},\C{k2}, and \C{k3}).  The body of
the loop executes a SQL update followed by an insert.  Recall that a
transaction reads from the global database ($\Delta$), and writes to a
transaction-local database ($\delta$) before committing these
updates. An update statement filters the records that match the search
criteria from $\Delta$ and computes the updated records that are to be
added to the local database. Thus, the state transformer for the
update statement (call it $\T_U$) is the following function on
sets\footnote{$\bind$ has higher precedence than $\cup$; angle braces
($\langle \ldots \rangle$) are used to denote records in math}:
\begin{smathpar}
\begin{array}{l}
  \lambda(\stl,\stg).~ \stl \cup \stg \bind(\lambda \C{s}. 
      \itel{{\sf table}(\C{s}) = \C{Stock} \conj 
        \C{s}.\C{s\_i\_id} = \C{item\_req.ol\_i\_id}\\\hspace*{1.15in}}
           {\{ \langle \C{s\_i\_id}=\C{s.s\_i\_id};\, 
                       \C{s\_d\_id}=\C{s.s\_d\_id};\,
                       \C{s\_qty} = \C{k1}\rangle \}\\\hspace*{1.15in}}
           {\emptyset})
\end{array}
\end{smathpar}
% \begin{smathpar}
% \begin{array}{c}
%   \lambda\Delta.\lambda\delta.\, \delta \;\cup\; 
%       \Delta \,\bind\, (\lambda s.\, 
%           \C{if}\;{\C{stock}(s) \conj 
%                    \C{s.s\_i\_id} = \C{item\_req.ol\_i\_id}}\\
%            \hspace*{0.9in}
%            \C{then}\;{\{\{s \;\C{with}\; \C{s\_qty} = \C{k1}\}\}}\;
%            \C{else}\;{\emptyset})
% \end{array}
% \end{smathpar}
% \begin{verbatim}
%   Rmem(σ'(s')) = Rmem(σ(s')) U 
%                  Rmem[\x. Rmem[\y. if y<x then {y+x} else {}](σ(s)) U {x}](xs)
% \end{verbatim}
% Observe that
% $\T_U(\Delta,\delta)$ is of the form $\delta \,\cup\,
% F_U(\Delta)$.
Here, the set bind operator extracts record elements ($\C{s}$) from
the database, checks the precondition of the update action, and if
satisfied, constructs a new set containing a single record that is
identical to $\C{s}$ except that it binds field $\C{s\_qty}$ to value
$\C{k_1}$.  This new set is added (via set union) to the existing
local database state $\delta$.\footnote{For now, assume that the
  record being added to $\stl$ is not already present in $\stl$.}
